---
title: "Introduction"
format: html
bibliography: countreg.bib
crossref:
  eq-prefix: ""
---

Modeling count variables is a common task in economics and the social
sciences. The classical Poisson regression model for count
data is often of limited use in these disciplines because empirical
count data sets typically exhibit over-dispersion and/or an excess
number of zeros. The former issue can be addressed by extending 
the plain Poisson regression model in various directions: e.g.,
using sandwich covariances or estimating an additional dispersion
parameter (in a so-called quasi-Poisson model). Another more formal
way is to use a negative binomial (NB) regression. All of these models
belong to the family of generalized linear models 
[GLMs, see @countreg:Nelder+Wedderburn:1972; @countreg:McCullagh+Nelder:1989].
However, although these models typically can capture over-dispersion
rather well, they are in many applications not sufficient for 
modeling excess zeros. Since @countreg:Mullahy:1986 and
@countreg:Lambert:1992 there is increased interest, both in the
econometrics and statistics literature, in zero-augmented models
that address this issue by a second model component capturing zero counts.
Hurdle models [@countreg:Mullahy:1986] combine a left-truncated
count component with a right-censored hurdle component. Zero-inflation
models [@countreg:Lambert:1992] take a somewhat different
approach: they are mixture models that combine a count component and a point mass at zero.
An overview of count data models in econometrics, including 
hurdle and zero-inflated models, is provided in
@countreg:Cameron+Trivedi:1998,countreg:Cameron+Trivedi:2005.

In R [@countreg:R:2008], GLMs are provided
by the model fitting functions `glm()` [@countreg:Chambers+Hastie:1992]
in the **stats** package and `glm.nb()` in the **MASS**
package [@countreg:Venables+Ripley:2002] along with associated
methods for diagnostics and inference. Here, we discuss the implementation
of hurdle and zero-inflated models in the functions `hurdle()` and
`zeroinfl()` in the **countreg** package [@countreg:Zeileis+Kleiber:2013],
available from the Comprehensive R Archive Network (CRAN)
at <http://CRAN.R-project.org/package=countreg>.
The design of both modeling functions as well as the methods operating
on the associated fitted model objects follows that of the base R
functionality so that the new software integrates easily into the
computational toolbox for modeling count data in R.

The remainder of this article provides an overview of the supported
count data models in **countreg**. More detailed discussions of the classical
count data models and their zero-augmented counterparts can be found in ["Basics"](basics.qmd) and ["Zero-augmented models"](zero-augmented.qmd),
respectively. Both links contain applications of the respective models to a microeconomic cross-section data set on the demand for medical care.


## Models and software {#sec-software}

```{r ovw-tab-cap}
#| echo: false

tab_cap <- paste("Overview of discussed count regression models.",
                 "All GLMs use the same log-linear mean function ($\\log(\\mu) = x^\\top \\beta$)",
                 "but make different assumptions about the remaining likelihood.",
                 "The zero-augmented models extend the mean function by",
                 "modifying (typically, increasing) the likelihood of zero counts.")
```

```{r ovw-tab}
#| echo: false
#| label: tbl-overview
#| tbl-cap: !expr tab_cap

model_type <- c("GLM", "", "", "", "zero-augmented", "")
distribution <- c("Poisson", "", "", "NB", "Poisson", "NB")
est_method <- c("ML", "quasi", "adjusted", "ML", "ML", "ML")
description <- c('Poisson regression: classical GLM, estimated by maximum likelihood (ML)',
                 paste('"quasi-Poisson regression": same mean function,',
                       'estimated by quasi-ML (QML) or equivalently generalized',
                       'estimating equations (GEE), inference adjustment via',
                       'estimated dispersion parameter'),
                 paste('"adjusted Poisson regression": same mean function,',
                       'estimated by QML/GEE, inference adjustment via',
                       'sandwich covariances'),
                 paste('NB regression: extended GLM, estimated by ML',
                       'including additional shape parameter'),
                 'zero-inflated Poisson (ZIP), hurdle Poisson',
                 'zero-inflated NB (ZINB), hurdle NB')

tab <- data.frame(Type = model_type, Distribution = distribution,
                  Method = est_method, Description = description)
tab <- tinytable::tt(tab)
tinytable::style_tt(tab, i = 4, line = "b", line_color = "lightgrey")
```

In this section, we briefly outline the theory and its implementation in 
R [@countreg:R:2008] for some basic count data regression
models as well as their zero-augmented extensions (see @tbl-overview
for an overview). The classical Poisson, geometric
and negative binomial models are described in a generalized linear model (GLM)
framework; they are implemented in R by the `glm()` function
[@countreg:Chambers+Hastie:1992] in the **stats** package and the
`glm.nb()` function in the **MASS** package [@countreg:Venables+Ripley:2002].
The hurdle and zero-inflated extensions of these models are provided by the
functions `hurdle()` and `zeroinfl()` in package **countreg** [@countreg:Zeileis+Kleiber:2013].
The original implementation of @countreg:Jackman:2008 was improved
by @countreg:Kleiber+Zeileis:2008 for **countreg** to make the fitting functions and the
fitted model objects more similar to their `glm()` and `glm.nb()` counterparts.
The most important features of the new `hurdle()` and `zeroinfl()` functions
are discussed below while some technical aspects are deferred to the appendix.

An alternative implementation of zero-inflated count models is available in
the currently orphaned package **zicounts**
[@countreg:Mwalili:2007]. Another extension of zero-inflated Poisson models is available
in package **ZIGP** [@countreg:Erhardt:2008] which allows dispersion---in addition to
mean and zero-inflation level---to depend on regressors.
However, the interfaces of both packages are less standard with fewer (or no)
standard methods provided. Therefore, re-using generic inference tools is more
cumbersome and hence these packages are not discussed here. 

Two packages that embed zero-inflated models into more general implementations
of GLMs and GAMs (generalized additive models) are **gamlss** 
[@countreg:Stasinopoulos+Rigby:2007] and **VGAM** [@countreg:Yee:2008].
The latter also provides hurdle models (under the name zero-altered models).
Both implementations allow specification of only one set of regressors.

In addition to zero-augmented models, there are many further extensions to
the classical Poisson model which are not discussed here. Some important model
classes include finite mixture models---implemented in R in package **flexmix**
[@countreg:Leisch:2004]---and generalized estimating equations
(GEE)---provided in R by package **geepack**
[@countreg:Halekoh+Hojsgaard+Yan:2006]---and mixed-effects models---available
in R in packages **lme4** and **nlme** [see @countreg:Pinheiro+Bates:2000].
Further information about the models and alternative R implementations
can be found in the respective references.
